---
title: "GIS Assignment 3"
author: "Marta Bratkowska, Luis Batista Tomé, Lyndsey Lewis, Jillian Hunter"
date: "2026-03-03"
output:
  html_document:
    toc: true
    number_sections: true
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load libraries used for both 
library(sf)
library(terra)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)

conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("lag", "dplyr")

knitr::opts_chunk$set(echo = TRUE)
# Q2
library(rnaturalearth)
library(rnaturalearthdata)
library(gdistance)
library(raster)
library(patchwork)

# Prefer dplyr's select() when packages mask it (prevents sf select() errors)
if (requireNamespace("conflicted", quietly = TRUE)) {
  conflicted::conflict_prefer("select", "dplyr", quiet = TRUE)
}
```



# Question 1
**Summary**
This Question shows climate change in the USA by average SPEI index across regions in the past 50 years. 

```{r pressure, echo=FALSE}
# Load file in
usstates <- spData::us_states
spei <- rast("dataQ1/spei01.nc")

# Aggregate based on the region column
usstates_aggr <- usstates %>%
  group_by(REGION) %>%
  summarise(geometry = st_union(geometry)) %>% 
  st_transform(crs = st_crs(spei))

test <- spei[[10]]
spat.crop <- crop(test, usstates_aggr)

# Though the plot in the assignment covers from 1965-2015, we generate a date vector for the past 50 years (using lubridate)
date_vector <- seq(from = as.Date("1974-01-16"), to = as.Date("2024-01-16"), by = "month")
# Numeric vector going from 781 to 1380
t <- time(spei)                 # time for every layer
idx_range <- 877:1477

numeric_vector <- tibble::tibble(
  spei_id = idx_range,
  Date = as.Date(t[idx_range])
)


```

```{r}
df <- result_test %>%
  select(877:1477)

df <- cbind(st_drop_geometry(usstates_aggr), df)


# excluding REGION, change all columns to long format
df_long <- df %>%
  pivot_longer(cols = -REGION, names_to = "spei_id", values_to = "value")

df_long <- df_long %>% 
  # using substr, remove first 10 characters from the "spei_id" column 
  mutate(spei_id = substr(spei_id, 11, nchar(spei_id))) %>% mutate(spei_id = as.numeric(spei_id))
  
# Join df_long with the numeric_vector (by "spei_id")
df_long <- df_long %>% 
  left_join(numeric_vector, by = "spei_id")

df_long <- df_long %>% 
  select(-spei_id) %>% 
  mutate(Date = ymd(Date))

# Group by year and summarize the mean value for each year and region
df_long_yearly <- df_long %>% 
  group_by(REGION, year = year(Date)) %>% 
  summarize(mean_value = mean(value, na.rm = TRUE)) %>% 
  ungroup()

```


```{r}
# Plot value over year for each REGION using ggplot2
ggplot(df_long_yearly, aes(x = year, y = mean_value, color = REGION)) +
  geom_line(linewidth = 0.7) + 
  geom_smooth(aes(group = 1), method = "loess", se = TRUE,
              color = "royalblue3", linewidth = 1) +
  labs(
    x = "year",
    y = "spei",
    color = "REGION"
  ) +
  theme_minimal()
```

# Question 2

**Summary**

This report solves Assignment 3: identify the 10 most populated places in mainland Spain, build a friction surface (roads cheap, off-road expensive), compute bilateral least-cost distances, and compare isolation for Madrid vs Vigo.

**Packages**



## 1. Load Natural Earth data

```{r data-download}
# Country polygon (10m)
countries <- ne_countries(scale = 10, returnclass = 'sf')
spain <- countries %>% filter(admin == 'Spain' | name == 'Spain')

# Populated places and roads (10m)
places <- ne_download(scale = 10, type = 'populated_places', category = 'cultural', returnclass = 'sf')
roads <- ne_download(scale = 10, type = 'roads', category = 'cultural', returnclass = 'sf')

# Ensure valid geometries
places <- st_make_valid(places)
roads <- st_make_valid(roads)

# Project to Spain metric CRS (UTM 30N)
crs_proj <- "EPSG:25830"
spain <- st_transform(spain, crs_proj)
places <- st_transform(places, crs_proj)
roads <- st_transform(roads, crs_proj)
```

## 2. Keep mainland Spain only & select top 10

```{r mainland-top10}
# Keep largest polygon (mainland, exclude islands)
spain_mainland <- spain %>%
  st_cast("MULTIPOLYGON") %>%
  st_cast("POLYGON") %>%
  mutate(area = as.numeric(st_area(.))) %>%
  arrange(desc(area)) %>%
  slice(1) %>%
  dplyr::select(-area)

# Filter places to mainland
places_mainland <- places %>%
  st_join(spain_mainland, left = FALSE)

# Detect population field robustly
cands <- c("POP_MAX","POP2020","POP2015","POP2000","POP_MIN")
pop_field <- cands[cands %in% names(places_mainland)][1]
if (is.na(pop_field)) stop("No population field found.")

places_mainland <- places_mainland %>%
  mutate(pop_est = as.numeric(.data[[pop_field]]))

# Select top 10
top10 <- places_mainland %>%
  arrange(desc(pop_est)) %>%
  slice(1:10)

cat("Top 10 populated places (mainland):\n", paste(top10$NAME, collapse=", "), "\n")
```

## 3. Build friction surface from mainland roads

```{r friction-raster}
# Clip roads to mainland Spain
roads_mainland <- st_intersection(roads, spain_mainland)

# Create raster template (1 km resolution) over mainland extent
res_m <- 1000
r_template <- rast(ext(vect(spain_mainland)), resolution = res_m, crs = crs_proj)

# Initialize with values (NA) before masking
values(r_template) <- NA
r_template <- mask(r_template, vect(spain_mainland))

# Rasterize roads: 1 where exists, NA elsewhere
roads_r <- rasterize(vect(roads_mainland), r_template, field = 1, background = NA)

# Create cost raster: roads cheap (1), off-road expensive (50), sea blocked (NA)
offroad_cost <- 50
road_cost    <- 1

r_cost <- r_template
values(r_cost) <- offroad_cost
values(r_cost)[!is.na(values(roads_r))] <- road_cost

# Final mask to mainland (ensure no sea movement)
r_cost <- mask(r_cost, vect(spain_mainland))

# Plot
plot(r_cost, main = 'Friction surface (mainland only)')
plot(st_geometry(spain_mainland), add = TRUE, border = 'black', lwd = 1.5)
plot(st_geometry(st_centroid(top10)), add = TRUE, col = 'red', pch = 20, cex = 1.5)
```

## 3b. Friction surface map (modern)

```{r friction-map, fig.height=6, fig.width=8}
# Convert raster to dataframe for ggplot
r_df <- as.data.frame(r_cost, xy = TRUE, na.rm = TRUE)
colnames(r_df)[3] <- "cost"

# Modern friction surface visualization
ggplot() +
  # Friction raster
  geom_raster(data = r_df,
              aes(x = x, y = y, fill = cost)) +
  
  # Mainland border
  geom_sf(data = spain_mainland,
          fill = NA,
          color = "black",
          linewidth = 0.4) +
  
  # Roads overlay (highlighted)
  geom_sf(data = roads_mainland,
          color = "white",
          linewidth = 0.35,
          alpha = 0.9) +
  
  # Modern color scale (roads bright, off-road darker)
  scale_fill_viridis_c(
    option = "magma",
    direction = -1,
    name = "Friction cost"
  ) +
  
  coord_sf() +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    panel.grid = element_blank()
  ) +
  labs(
    title = "Friction Surface – Mainland Spain",
    subtitle = "Roads (low cost) vs Off-road (high cost)"
  )
```

## 4. Compute 10×10 bilateral least-cost distance matrix

```{r gdistance}
# Convert terra SpatRaster to raster::Raster for gdistance
r_cost_raster <- raster::raster(r_cost)

# Build transition object: conductance = 1/cost
tr <- transition(
  x = r_cost_raster,
  transitionFunction = function(x) 1 / mean(x),
  directions = 8
)
tr <- geoCorrection(tr, type = "c")

# Convert top10 to Spatial for gdistance
top10_sp <- as(top10, "Spatial")

# Compute FULL 10x10 bilateral matrix
dist_mat <- costDistance(tr, top10_sp) %>% as.matrix()

# Label rows/columns
city_names <- top10$NAME
rownames(dist_mat) <- city_names
colnames(dist_mat) <- city_names

# Display matrix
knitr::kable(dist_mat, caption = "Least-cost distance matrix (10×10, friction units)")
```

## 5. Compare isolation: Madrid vs Vigo

```{r madrid-vigo}
# Extract distance vectors for Madrid and Vigo
madrid_vec <- dist_mat["Madrid", setdiff(city_names, "Madrid")]
vigo_vec   <- dist_mat["Vigo",   setdiff(city_names, "Vigo")]

# Summary statistics
iso_tbl <- data.frame(
  Origin = c("Madrid", "Vigo"),
  Mean = c(mean(madrid_vec), mean(vigo_vec)),
  Median = c(median(madrid_vec), median(vigo_vec)),
  Max = c(max(madrid_vec), max(vigo_vec))
)
rownames(iso_tbl) <- NULL

knitr::kable(iso_tbl, digits = 0, caption = "Isolation summary: bilateral least-cost distances (friction units)")

# Determine which is more isolated
madrid_mean <- mean(madrid_vec)
vigo_mean <- mean(vigo_vec)

if (madrid_mean > vigo_mean) {
  isolation_result <- paste0("Madrid is more isolated (mean: ", round(madrid_mean, 0), 
                             " vs Vigo: ", round(vigo_mean, 0), " friction units).")
} else if (vigo_mean > madrid_mean) {
  isolation_result <- paste0("Vigo is more isolated (mean: ", round(vigo_mean, 0), 
                             " vs Madrid: ", round(madrid_mean, 0), " friction units).")
} else {
  isolation_result <- "Madrid and Vigo have equal isolation."
}

cat("**Result:**", isolation_result)
```

## 5b. Validation: Madrid → Vigo least-cost path

```{r shortest-path, fig.height=5, fig.width=6}
# Compute shortest (least-cost) path Madrid → Vigo
madrid_cent <- st_centroid(top10 %>% filter(NAME == "Madrid"))
vigo_cent   <- st_centroid(top10 %>% filter(NAME == "Vigo"))

orig_coord <- st_coordinates(madrid_cent)[1, ]
dest_coord <- st_coordinates(vigo_cent)[1, ]

path_mv <- shortestPath(tr, origin = orig_coord, goal = dest_coord, output = "SpatialLines")
path_mv_sf <- st_as_sf(path_mv) %>% st_set_crs(crs_proj)

# Visualize
ggplot() +
  geom_sf(data = spain_mainland, fill = 'lightgrey', color = 'black', linewidth = 0.3) +
  geom_sf(data = roads_mainland, color = 'grey50', linewidth = 0.2, alpha = 0.6) +
  geom_sf(data = st_centroid(top10), color = 'red', size = 2) +
  geom_sf(data = path_mv_sf, color = 'blue', linewidth = 1.2) +
  theme_minimal() +
  labs(title = 'Least-cost path: Madrid → Vigo (blue line)',
       subtitle = 'Path follows road corridors (validates friction model)')
```

## 5c. Density comparison: Madrid vs Vigo

```{r madrid-vigo-density, fig.height=5, fig.width=8}
# Prepare data frame for density plot
df_long <- bind_rows(
  data.frame(origin = "Madrid", value = as.numeric(madrid_vec)),
  data.frame(origin = "Vigo",   value = as.numeric(vigo_vec))
)

# Plot overlapping distributions
ggplot(df_long, aes(x = value, fill = origin)) +
  geom_density(alpha = 0.4, color = NA) +
  scale_fill_manual(values = c('Madrid' = '#F8766D', 'Vigo' = '#00BFC4')) +
  theme_minimal() +
  labs(
    x = 'Least-cost distance (friction units)',
    y = 'Density',
    title = 'Bilateral distance distributions: Madrid vs Vigo',
    fill = 'City'
  ) +
  theme(legend.position = 'right')
```

## 6. Results and conclusions

**Assignment objectives achieved:**

- ✓ **Mainland Spain only**: Islands excluded via polygon filtering
- ✓ **Top-10 populated places**: `r paste(top10$NAME, collapse = ', ')`
- ✓ **Friction surface**: Roads (cost=1) vs. off-road (cost=50), masked to mainland
- ✓ **10×10 bilateral matrix**: Full least-cost distance matrix computed
- ✓ **Madrid vs Vigo**: `r isolation_result`
- ✓ **Validation**: Shortest path plotted (confirms road-following behavior)

**Summary statistics:**
- Madrid mean isolation: `r round(madrid_mean, 0)` friction units
- Vigo mean isolation: `r round(vigo_mean, 0)` friction units
- Range of distances: `r round(min(dist_mat[dist_mat > 0], na.rm=TRUE), 0)` – `r round(max(dist_mat, na.rm=TRUE), 0)` units

---

## Appendix: Save outputs

```{r save-outputs, echo=FALSE}
# Save full distance matrix
write.csv(dist_mat, file = 'pairwise_cost_distances_top10.csv')
cat("✓ Pairwise distance matrix saved to: pairwise_cost_distances_top10.csv\n")

# Save summary table
write.csv(iso_tbl, file = 'isolation_summary.csv', row.names = FALSE)
cat("✓ Isolation summary saved to: isolation_summary.csv\n")
```

**Notes:**
- Distances are in **friction units** (not calibrated to meters/km); based on cost surface (road=1, offroad=50).
- If data download fails: replace `ne_download()` calls with local shapefiles.
- Knit command: `rmarkdown::render('assignment3.Rmd')` in R console.
